Mining HIV Trends in Social Media Data

### 2016-04-19 draft/draft.tex
work more on draft, notebooks still not quite fixed, actually

Here is the abstract.

@@

@@

@@

We excluded non-English tweets, and also tweets that did not originate in Contiguous United States time zones. A subset of the tweets we collected had geolocation data available. We found that tweets were largely concentrated on US and Canadian metro areas (Figure 1).

@@

@@

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to 'truvada'}
\begin{tabular}{|l|c|} \hline
Related Word & Cosine Similarity to 'truvada'\\ \hline
truvada & 0.796666\\ \hline
Truvada & 0.793999\\ \hline
DoingIt & 0.738141\\ \hline
WorldAidsDay & 0.728808\\ \hline
WorldAIDSDay & 0.720910\\ \hline
NancyReagan & 0.717667\\ \hline
NBHAAD & 0.705061\\ \hline
ART & 0.704300\\ \hline
hiv & 0.702844\\ \hline
HLM2016AIDS & 0.698698\\ \hline
\hline\end{tabular}
\end{table}

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to '\#truvada'}
\begin{tabular}{|l|c|} \hline
Related Hashtag/Tweet & Cosine Similarity to '\#truvada'\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 0.727602\\ \hline
\#whereisprep & 0.707165\\ \hline
\#truvada & 0.696113\\ \hline
\#hivprevention & 0.636068\\ \hline
702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
708519265540907010 & 0.628778\\ \hline
712032637024653313 & 0.628646\\ \hline
\#harrogatehour & 	0.628547\\ \hline
\hline\end{tabular}
\end{table}

@@

% PrEP related trends (2 topic time series worth)
% World AIDs day as validation (1 topic time series)

Next we sought to identify some temporal trends in PrEP related trends. We used Dynamic Topic Modelling to identify how certain topics change over time. We specified 10 topics and plotted representative words from two of the topics.

@@

In topic 3 we see that the keyword 'prep' remains relatively constant, while 'prevention' and 'truvada' decline over time. 'hivaids' increases and then decreases in magnitude over the course of the time series (Figure 2). These dynamics may indicate short term events in hivaids and PrEP-related conversation on twitter over the time period studied.

We found that topic 4 captured several keywords related to World AIDS Day (Figure 3). We can see that all of these terms peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore,while December 1st is World AIDS day, the whole month of Decemeber is AIDS Awarness month. We are clearly seeing the word 'awarness' peak later and last longer that the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While we aren't specifically interested in World AIDS Day, or AIDS awareness month, to inform our understanding of PrEP discussion, this observation validates our ability to identify temporal events using DTM.

% breifly mention what was in the other topics (lots of AIDS related news announcements)

% figure 2
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig1}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

% figure 3
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig2}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

@@

 this will take time to complete (running download, prob 5 days, so approx April 19th)

1 figure (heatmap of users vs topics)

1 table (top words for 10 topics)

@@

% top positive and negative example tweets
% report accuracy
% (potentially) report topic modeling on top postive and negative PrEP tweets

Lastly we sought to create a classification scheme to classify the sentiment of tweets either positive or negative. This classifier would allow us to quickly identify positive and negative PrEP related tweets to guide public health efforts. We first performed Paragraph2Vec on both our PrEP related tweet corpus and on another tweet corpus that had binary sentiment labels, positive, or negative. Then we trained a simple logistic regression classifier on the paragraph-vectors. We found that our classifier had an accuracy of 69\% on a validation set.

@@

Conclusions go here.

>>>>

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV). The treatment is available as a pill named Truvada and is taken once a day for as much as 99\% protection from infection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. This paper contributes a data mining and machine learning strategy that monitors Twitter social media data. Our results show that we can identify trends over time, and positive and negative PrEP related tweets. These results can be used by public health professionals to inform and protect at risk individuals from HIV.

%These methods may also be useful to identify and preempt HIV outbreaks before they reach epidemic levels.

@@

 - High cost of actual drug medication, and unclear health insurance situation.

 - Many health providers do not know about drug.

@@

 - Here is a paper that did twitter data mining for flu prediction:

 %http://snap.stanford.edu/soma2010/papers/soma2010_16.pdf

 Talk about how it's harder to get feedback/ account for HIV infections compared with CDC baseline data for flu.

@@

We excluded non-English tweets, and also tweets that did not originate in contiguous United States time zones. A subset of the tweets we collected had geolocation data available. We found that tweets were largely concentrated on US and Canadian metro areas (Figure 1).

@@

% TODO work this section Tuesday

@@

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to 'PrEP'}
\begin{tabular}{|l|c|} \hline
Related Word & Cosine Similarity to 'PrEP'\\ \hline
\hline\end{tabular}
\end{table}

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to '\#truvada'}
\begin{tabular}{|l|c|} \hline
Related Hashtag/Tweet & Cosine Similarity to '\#truvada'\\ \hline
\hline\end{tabular}
\end{table}

@@

Next we sought to identify some temporal trends in PrEP related trends. We used Dynamic Topic Modelling (DTM) to identify how certain topics change over time. In DTM, documents are grouped in several corpuses that represent different points in time. Latent Dirichlet Analysis (LDA) is then performed on each corpus, to extract a topic distribution. The posterior distribution from time $t_n$ is used as the prior for time $t_{n+1}$. We specified 10 topics and used each week's worth of tweets over a 30 week period as our time points.

@@

We identified two topics that showed relevant trends. In topic 5 we see that the keyword 'prep' remains relatively constant, while 'prevention' and 'truvada' decline over time. 'hivaids' increases and then decreases in magnitude over the course of the time series (Figure 3). These dynamics may indicate short term events in hivaids and PrEP-related conversation on twitter over the time period studied.

% figure 3
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig1}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

We found that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that all of these terms peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore,while December 1st is World AIDS day, the whole month of December is AIDS Awareness month. We are clearly seeing the word 'awareness' peak later and last longer that the word 'WorldAIDSDay' implying that it is correlated to the whole month of December. While we aren't specifically interested in World AIDS Day, or AIDS awareness month, to inform our understanding of PrEP discussion, this observation validates our ability to identify temporal events using DTM.

% figure 4
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{DTMfig2}
\caption{DTM topic 5, word prevalence over time. Date is YYYY-WW.}
\end{figure}

These results demonstrate our ability to extract relevant HIV and PrEP related information from twitter that accurately captures time-dependant fluctuations. Public health professionals should be able to monitor these temporal trends to determine the relative interest in PrEP related social media discourse. This could lead to a system for early detection of HIV outbreak detection, or other serious developments.

@@

% TODO work this section (re download and re-run) Tuesday

We wanted to identify what Twitter users that mentioned PrEP were discussing in their other tweets. We identified users that were most similar to PrEP according to our Paragraph2Vec results, and downloaded their most recent 3000 tweets. We took the top 300 users that had at least 200 words in the combined tweets of their tweet history. For each user, we concatenated all of their tweets, and performed LDA topic modelling on the set of user-time line documents.

When we examine a clustermap of users vs. topics, we see that the majority of these users are discussing topics 5, 1, and 4 (Figure 5). When we look up the top words in the corresponding topic distributions, we see that topics 1 and 4 seem to be related to republican party politics.

% figure 5
\begin{figure}
\centering
\includegraphics[height=2.5in, width=3.5in]{user_timeline_clustermap}
\caption{Clustermap of users vs. topics for the top 300 users related to PrEP.}
\end{figure}

% table 3
\begin{table*}
\centering
\caption{Topic modeling on PrEP-related users\' timelines.}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|} \hline
Topic Number & Word 0 & Word 1 & Word 2 & Word 3 & Word 4 & Word 5 & Word 6 & Word 7 & Word 8 & Word 9\\ \hline
0 & hivaids & condoms & bitcoin & stigma & cdcstd & kanyewest & nyhaad & entrepreneur & realestate & blacklivesmatter\\ \hline
1 & ang & ng & tcot & ako & sa & ko & pjnet & realdonaldtrump & gopdebate & cruzcrew\\ \hline
2 & fertility & infertility & meme & mxico & rp & deactivate & rail & roc & offensive & feedly\\ \hline
3 & colombia & pozlivingbc & fe & repostapp & sprite & bccfe & thewalkingdead & npinchat & interior & wellness\\ \hline
4 & businessinsider & stacks & noh & teaparty & sk & holi & alzheimers & caregiver & asa & guru\\ \hline
5 & armenia & wordpress & georgia & topping & lgbtqia & yah & faction & checkout & narendramodi & examination\\ \hline
6 & knitting & lista & naral & gust & worldbank & bigdata & haiku & vdeo & mhealth & hah\\ \hline
7 & nigga & niggas & bernie & lmao & nhsengland & bitches & bae & uniteblue & il & gop\\ \hline
8 & socialmedia & seo & socialmediamarketing & der & und & entrepreneur & ist & nowplaying & foxbusiness & enews\\ \hline
9 & collage & mileycyrus & undetectable & ddlovato & casa & cnet & qotd & vacations & nashville & heros \\ \hline

\hline\end{tabular}
\end{table*}

Together these results suggest that PrEP users are involved in political discussions, and are internationally focused.

@@

% TODO work this section Tuesday (give example positive and negative tweets)

Lastly we trained a classifier to classify the sentiment of HIV and PrEP related tweets either positive or negative. This classifier would allow public health professionals to quickly identify positive and negative PrEP related tweets to guide HIV prevention efforts. We obtained a set of 1.6 million tweets with sentiment labels, either positive or negative from Sanders Analytics (http://www.sananalytics.com/lab/twitter-sentiment/). Then we trained a simple logistic regression classifier on 1.2 million paragraph-vectors from the sentiment dataset. We found that our classifier had an accuracy of 69\% using a portion of the sentiment tweets not used in training, as a validation set.

 We chose to use a relatively simple classifier model (logistic regression) and stop training at

@@

- Moving towards HIV outbreak prediction.

- Automatically identifying PrEP adoption/adherence sentiment in US population.
